---
title: "Figure 2: Characterization of spatial variation and patterning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document, we reproduce panels from Figure 2 of our cerebellar cortex transcriptomic atlas manuscript. 

## Load utility functions and data

```{r load, results='hide', message=F, warning=F}
library(Seurat)
# v2.3.4
library(ggplot2)

# Utility functions
source('../src/fig2_utils.R')

```

We first load annotated adult mouse cerebellar cortex data. We can retrieve this in raw count format from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE165371) (cb_adult_mouse tar). Download and expand this directory and copy the files into the `data` directory of this repo. 

``` {r data, eval=F}
# Read in full dataset (as mtx or RDS file)
# Raw counts can be downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE165371
# (cb_adult_mouse)
# Place in data directory
counts_cb = readMM('../data/cb_adult_mouse.mtx.gz')
barcodes_cb = readLines('../data/cb_adult_mouse_barcodes.txt')
genes_cb = readLines('../data/cb_adult_mouse_genes.txt')

row.names(counts_cb) = genes_cb
colnames(counts_cb) = barcodes_cb

# Create Seurat object if reading in raw data
full_cb = CreateSeuratObject(raw.data = counts_cb)
```

Alternatively, we can download a Seurat object with UMAP representation stored directly from the Single Cell Portal [study](https://singlecell.broadinstitute.org/single_cell/study/SCP795). Download the `cb_annotated_object.RDS` object from the Download tab of the study and copy into the `data` directory. (Note that you will have to sign into SCP in order to access the Download tab.)

``` {r data_obj}
# Seurat object can be downloaded from SCP
# place in data directory
full_cb = readRDS('../data/cb_annotated_object.RDS')
```

We add in the cluster and subcluster annotations. This also includes additional metadata like sample sex. (Note that the SCP object also contains these annotations.)

```{r metadata}
# read in cluster metadata (available in data directory)
cluster_metadata = read.csv('../data/cluster_metadata.csv', header = T, row.names = 1)
full_cb@meta.data = cluster_metadata

# also read in metadata corresponding to object with granule cells downsampled to 60k
cluster_metadata_ds = read.csv('../data/cluster_metadata_ds.csv', header = T, row.names = 1,
                               stringsAsFactors = F)
```

```{r preprocess}
# split up by analyses 
cluster_metadata_ds[['analysis_cluster']] = sapply(cluster_metadata_ds$cluster, function(x) {
  if (x %in% c('OPC', 'ODC')) {
    return('OPC/ODC')
  } else if (x %in% c('PLI', 'MLI1', 'MLI2')) {
    return('MLI/PLI')
  } else if (x %in% c('Macrophages', 'Microglia')){
    return('Microglia')
  } else if (x %in% c('Choroid', 'Ependymal')) {
    return('Choroid')
  } else if (x %in% c('Dcn_Fibroblasts', 'Endothelial_mural', 'Endothelial_stalk')) {
    return('Endothelial/Fibroblasts')
  } else {
    return(x)
  }
})

cluster_metadata_ds[['lob_comp_cluster']] = sapply(cluster_metadata_ds$subcluster, function(x) {
  if (x %in% c('PLI_1', 'PLI_2', 'PLI_3')) {
    return('PLI')
  } else {
    return(x)
  }
})
cluster_metadata_ds$lob_comp_cluster = factor(cluster_metadata_ds$lob_comp_cluster)
cluster_metadata_ds$analysis_cluster = factor(cluster_metadata_ds$analysis_cluster)
```

```{r polygons}
# Prepare cerebellar cartoon polygons for plotting later panels
# Lobule coordinates available in data directory
coords_full = c('coordsI_half.csv',  'coordsII_half.csv', 'coordsIII_half.csv', 
                'coordscul_half.csv', 'coordsVI_half.csv', 'coordsVII_half.csv',
                'coordsVIII_half.csv', 'coordsIX_half.csv', 'coordsX_half.csv',
                'coordsan12.csv', 'coordsan22.csv', 'coordsprm2.csv', 
                'coordssim2.csv', 'coordscop2.csv', 'coordsf2.csv', 
                'coordspf2.csv')

coords_full = paste0('../data/coords/', coords_full)

coords_names = c('I', 'II', 'III', 'CUL', 'VI', 'VII', 'VIII', 'IX', 'X',
                 'AN1', 'AN2', 'PRM', 'SIM', 'COP', 'F', 'PF')

SPs = generate_SPs(coords_full, coords_names)

```

## Panel A

Spatial divergence computation and plotting.

``` {r compute}
celltypes_test = c('Purkinje', 'Granule', 'OPC/ODC', 'Golgi', 'UBC',
                   'MLI/PLI', 'Bergmann', 'Astrocyte' )

for (celltype in celltypes_test) {
  celltype_sub = cluster_metadata_ds[cluster_metadata_ds$analysis_cluster == celltype,]
  celltype_sub$lob_comp_cluster = droplevels(celltype_sub$lob_comp_cluster)
  celltype_df = getScatterPlotData_meta(celltype_sub, celltype = celltype)
  
  if (exists('full_scatter_df')) {
    full_scatter_df = rbind(full_scatter_df, celltype_df)
  } else {
    full_scatter_df = celltype_df
  }
}

full_scatter_df[['log_fc']] = log(full_scatter_df$fc_values, base = 2)
full_scatter_df[['log_p_values']] = -1 * log(full_scatter_df$p_values + 1e-250, base = 10)
full_scatter_df[['log_q_values']] = -1 * log(full_scatter_df$q_values + 1e-250, base = 10)
```

```{r scatterploto}
ggplot(data = full_scatter_df, aes(x = log_q_values, y = log_fc)) + 
  geom_point(data = full_scatter_df, aes(color = celltype), size = 2) +
  xlab('-log10(q-value)') +
  ylab('log(Max Region FC)') + theme(legend.title = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size=4))) + 
  geom_vline(xintercept = 3, linetype = 'dashed', col = 'grey')
```

## Panel B (Purkinje cells)

```{r dotplot_purk, results='hide'}
# Subset to purkinjes
full_cb = SetIdent(full_cb, ident.use = full_cb@meta.data$subcluster)
purk_s = SubsetData(full_cb, subset.name = 'cluster', accept.value = 'Purkinje', 
                    subset.raw = T)

#  remember to normalize data
purk_s = NormalizeData(purk_s)
genes_purkinje = rev(c("Aldoc",'Kctd16', 'Gpr176', 'Drd3', "Ephb2","Tshz2","Fam19a2",
                       "Alcam","Tox2","2900055J20Rik","Vmn2r30"))

# pdf("Figure2_panelB.pdf", useDingbats = F, width = 5, height = 3.5)
# use slightly customized dotplot function
DotPlot2(purk_s, genes_purkinje,
         scale.by = 'radius',dot.min = 0.2, dot.scale = 6,
         scale.min = 0.1,plot.legend = T, x.lab.rot = 45)
# dev.off()
```

## Panel C (Aldoc+ and - distributions)

```{r aldoc_map}

# split into aldoc+ and aldoc-
purk_s@meta.data[['Aldoc_stat']] = sapply(as.character(purk_s@ident), function(x) {
  split = strsplit(x, '_')[[1]]
  if (split[2] == 'Anti') {
    return('Aldoc_neg')
  } else {
    return('Aldoc_pos')
  }
})

purk_s@meta.data$Aldoc_stat = factor(purk_s@meta.data$Aldoc_stat)
# plot spatial distributions
# pdf('Figure2_PanelC_left.pdf', useDingbats = F)
spatial_enrichment_map_meta(purk_s@meta.data, meta_col = 'Aldoc_stat', sps = SPs, 
                            cluster = 'Aldoc_pos', do.print = F, color_divergent = T)
# dev.off()

# pdf('Figure2_PanelC_right.pdf', useDingbats = F)
spatial_enrichment_map_meta(purk_s@meta.data, meta_col = 'Aldoc_stat', sps = SPs,
                            cluster = 'Aldoc_neg', do.print = F, color_divergent = T)
# dev.off()
```

## Panel D (Purkinje subcluster distributions)

```{r purk_map}

# plot spatial distributions of subclusters
# pdf('Figure2_panelC.pdf', useDingbats = F)
# spatial distributions for all purkinje clusters 
for (clust in levels(purk_s@ident)) {
  print(spatial_enrichment_map_meta(purk_s@meta.data, meta_col = 'subcluster', sps = SPs,
                                    cluster = clust, do.print = F, color_divergent = T))
}

# dev.off()

```

## Panel E (Granule cells)

```{r gran_dotplot, fig.width=4.5, fig.height=3, results='hide'}
# clean up
rm(purk_s)
gc()

gran_s = SubsetData(full_cb, subset.name = 'cluster', accept.value = 'Granule',
                    subset.raw = T)

gran_s = NormalizeData(gran_s)
genes_gran = c("Chrm3","Nr4a2","Galntl6", 'Ebf1', "Gprin3", "Rasgrf1")

# pdf("Figure2_panelE.pdf",useDingbats = F, width = 5, height = 3.5)
DotPlot2(gran_s, genes_gran, dot.scale = 10,
         scale.min = 0.1, plot.legend = T,
         x.lab.rot = 45)
# dev.off()

```

## Panel F (Granule subcluster distributions)

```{r gran_map}

# plot spatial distributions of subclusters

# pdf('Figure2_panelF.pdf', useDingbats = F)
# spatial distributions for all granule clusters 
for (clust_i in 1:3) {
  clust = levels(gran_s@ident)[clust_i]
  print(spatial_enrichment_map_meta(gran_s@meta.data, meta_col = 'subcluster', sps = SPs,
                                    cluster = clust, do.print = F, color_divergent = T))
}

# dev.off()

```
## Panel G (Bergmann glia)

```{r berg_dotplot, fig.width=4.5, fig.height=3, results='hide'}
# clean up
rm(gran_s)
gc()

bergmann_s = SubsetData(full_cb, subset.name = 'cluster', accept.value = 'Bergmann',
                    subset.raw = T)

bergmann_s = NormalizeData(bergmann_s)

# pdf('Figure2_panelG', useDingbats = F, height = 3, width = 4)
DotPlot2(bergmann_s, c('Wif1', 'Mybpc1'), dot.scale = 8,
         scale.min = 0.1, plot.legend = T, scale.by = 'size',
         x.lab.rot = 45)
# dev.off()

```

## Panel H (Bergmann glia subcluster distributions)

```{r berg_map}

# plot spatial distributions of subclusters

# pdf('Figure2_panelH.pdf', useDingbats = F)
# spatial distributions for Bergmann cluster 
spatial_enrichment_map_meta(bergmann_s@meta.data, cluster = 'Bergmann_2', meta_col = 'subcluster', sps = SPs,
                       do.print = F, color_divergent = T)

# dev.off()

```


```{r packages}
sessionInfo()
```

